\documentclass[12pt]{article}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{fancyhdr}
\usepackage{mathptmx}
\usepackage[letterpaper, portrait, margin=1in,headheight=60.pt]{geometry}

\title{}
\begin{document}

\fontfamily{ptm}
% Outline: 
% Why is the regularity question in Navier Stokes important?
% 	Terry Tao introduction
% 	Step back to simpler case: Euler equations 
% How do we hope to find blowup?
% 	Axis-symmetric flow might lead to blowup
%		Tornado and tea cup example
%	Perhaps we can create blowup at the boundary
% What kind of model do we choose?
%	Need to close the system
%		Look for proper biot-savart law
% Why our model is special?
%	Better model for 3D Euler	
% How did we solve the problem and how will we approach the next problem?
%	Solved it via numerics and theory
% Why do our results matter?
%	 
% Future directions:
%	Can the blowup for our equation be translated into blowup for Euler?
%	
% Questions for Sverak:
% 1. How could we use this model to generate blowup for Euler 3D?
% 2. If we can rule out such a singularity near the boundary for Navier Stokes, then what is the point of this?
% 3. Why do we flip the velocity field in the initial configuration?
% 4. Why does he expect a singularity at the center point
% 5. Why don't we care about Burger's equation?
% 6. Is there something special about the 1D equation that we hope will connect the blowup to the 2D equation?
% 7. Why isn't there a canonical choice?

During his interview on the Colbert Report, Professor Terrance Tao told the audience that he was ``trying to see if water can spontaneously blow up.'' Looking concerned, Colbert interrupted and said ``Terrance, that would be \textit{good to know}.'' Indeed, existence of smooth initial data that leads to blowup for the Navier-Stokes equations is a crucial problem in fluid dynamics. It is also fascinatingly difficult.

A more tractable problem might be to find blowup solutions to Eulers' equations, that is, restrict attention to inviscid fluids (fluids that behave like water). But even two- and three-dimensional Eulers' equations are difficult to attack in generality. Hence, to study only a one-dimensional flow, we can consider the boundary for sufficiently symmetric flows.

Following a thought experiment of Einstein [1], imagine a teacup with tea-leaves at the bottom. Stirring with a spoon generates a cylindrical flow. Does this configuration generate a singularity? The answer is in the tea-leaves: does the flow sweep the leaves to the wall of the mug or two the center? Surprisingly, the swirl pushes the leaves to the center of the bottom of the mug and this point is where a singularity might lie.

This center point might have a singularity because the stirring motion generates a secondary-flow as shown that sweeps the tea-leaves inwards. Due to the opposing swirls, one expects the vorticity to be quite large, possibly large enough to produce blowup. More importantly, this point lies on the boundary, which due to the symmetry of the flow, we can assume is one dimensional.

We need to decide how to model this one dimensional flow. The are two choices: how to transport vorticity, and how to relate vorticity to velocity (Biot-Savart law). Simply reducing Eulers' equations to one-dimension gives Burgers' equation
\[
	u_t + u u_x = 0	
\]
which exhibits shocks, but not blowup [cite]. Yudovich [2] studied an analogy to two-dimensional Euler given by
\[
	\omega_t + u \omega_x = 0, \quad u_x = H \omega
\]
but showed that no blowup occurs. De Gregorio [3] proposed an analogy to three-dimensional Euler given by
\[
	\omega_t + u \omega_x = u_x \omega, \quad u_x = H \omega,
\]
or, written with the Lie bracket, as
\[
	\omega_t + [u, \omega] = 0.
\]
Surprisingly, despite attempts [4] to settle the question, global existence and regularity has remained open.

(paragraph about our results)

Having established regularity,



\end{document}
